Causes and effects in Dichotomous Comparative Judgments: an information-theoretical system of plausible mechanism

### 2024-08-16 _tex/index.tex
Built site for gh-pages

This section introduces fundamental concepts in causal inference, such
as directed acyclic graphs (DAGs), structural causal models (SCMs), and
the flow of association and causation in graphs. The section is not a
comprehensive description of causal inference methods. Readers
interested in deeper exploration should consult introductory papers like
\citet{Pearl_2010}, \citet{Rohrer_2018}, \citet{Pearl_2019}, and
\citet{Cinelli_et_al_2020}. They may also find introductory books such
as \citet{Pearl_et_al_2018}, \citet{Neal_2020} and
\citet{McElreath_2020} useful. For more advanced study, seminal
intermediate papers like \citet{Neyman_et_al_1923}, \citet{Rubin_1974},
\citet{Spirtes_et_al_1991}, and \citet{Sekhon_2009}, as well as books
like \citet{Pearl_2009}, \citet{Morgan_et_al_2014} and
\citet{Hernan_et_al_2020} are recommended.

@@

Empirical research addresses real-world challenges by relying on
evidence gathered through observation and experimentation. In this
context, researchers typically frame their research questions as
\emph{estimands} or \emph{targets of inference}. These estimands
represent the specific quantities the study aims to determine
\citep{Everitt_et_al_2010}. For instance, a study might examine the
question, ``To what extent do different teaching methods (\(X\))
influence students' conceptual understanding of a topic (\(Y\))?'' To
investigate this, the study could randomly assign students to two
groups, each using a different teaching method \((X=\{1,2\})\).
Students' conceptual understanding of the topic could be evaluated
through pairwise comparisons, resulting in a dichotomous outcome
\((Y=\{0,1\})\), indicating which student among those compared has a
higher level of understanding. The research question could be then
framed as the estimand, ``\emph{On average}, is there a difference in
conceptual understanding of the topic between the two groups of
students?'' This estimand could be mathematically expressed as
\(E[Y| X=1] - E[Y| X=2]\), where \(E[\cdot]\) denotes the expected
value. An example of this approach is seen in \citet{Jones_et_al_2019}.

Researchers then proceed to identify the estimands.
\emph{Identification} refers to the process of accurately computing an
estimand using an estimator. An \emph{estimator} is a method or function
that maps data into an estimate \citep{Neal_2020}. \emph{Estimates} are
numerical values that approximate the estimand and are derived through
\emph{estimation}, which refer to the process of integrating data with
an estimator \citep{Everitt_et_al_2010}. Although various methods can
approximate an estimand, researchers prioritize estimators with
desirable properties that ensure the accuracy of estimates. For
instance, the BTL model \citep{Bradley_et_al_1952, Luce_1959} is an
estimator known for its effectiveness in modeling pairwise comparisons,
yielding accurate estimates for the target of inference discussed
earlier, as long as its assumptions are met.

However, many studies aim to understand the mechanisms underlying
specific data types and also seek to establish causal relationships
rather than merely infer associations. In the earlier example, the
associational estimates obtained using the BTL model can be interpreted
as causal, because the data were collected using a randomized
experiment. Randomized experiments enable the causal interpretation of
associational estimates by ensuring several key properties, including
common support, no interference, and consistency
\citep{Morgan_et_al_2014, Neal_2020}. The most crucial property,
however, is that randomization effectively eliminates confounding.
\emph{Confounding} occurs when an external variable influences both the
outcome and the variable of interest, leading to spurious association
\citep{Everitt_et_al_2010}. Randomization mitigates this issue by
decoupling the intervention assignment mechanism, such as assigning
students to different groups, from other variables and outcomes
\citep{Morgan_et_al_2014, Neal_2020}.

@@

Although a detailed explanation of causes, and counterfactuals is beyond
the scope of this document, it suffices to say that causal inference
addresses the \emph{fundamental problem of causality} through the use of
counterfactuals \citep{Neal_2020}. The framework defines the
\emph{individual causal effect} (ICE) as the difference between an
individual's observed and unobserved potential outcomes. These
unobserved potential outcomes, termed \emph{counterfactuals}, represent
hypothetical scenarios that are \emph{contrary to fact}, where
alternative outcomes for an individual, resulting from a specific cause,
are neither observed nor observable
\citep{Neal_2020, Counterfactual_2024}. The framework then extends this
estimand to groups by introducing the \emph{average causal effect}
(ACE). The ACE reflects the difference between averages of observed
potential outcomes and counterfactuals across groups. Finally, similar
to an experiment, the framework identifies the ACE from associational
estimates by ensuring the absence of confounding.

@@

@@

\subsection{A motivating example}\label{sec-framework-example}

>>>>

This section introduces fundamental concepts in causal inference but
does not offer a comprehensive description of causal inference methods.
Readers interested in deeper exploration should consult introductory
papers like \citet{Pearl_2010}, \citet{Rohrer_2018}, \citet{Pearl_2019},
and \citet{Cinelli_et_al_2020}. They may also find introductory books
such as \citet{Pearl_et_al_2018}, \citet{Neal_2020} and
\citet{McElreath_2020} useful. For more advanced study, seminal
intermediate papers like \citet{Neyman_et_al_1923}, \citet{Rubin_1974},
\citet{Spirtes_et_al_1991}, and \citet{Sekhon_2009}, as well as books
like \citet{Pearl_2009}, \citet{Morgan_et_al_2014} and
\citet{Hernan_et_al_2020} are recommended.

@@

Empirical research addresses real-world challenges by relying on
evidence gathered through observation and experimentation. In this
context, researchers typically frame their research questions as
\emph{estimands} or \emph{targets of inference}. These estimands
represent the specific quantities the study aims to determine
\citep{Everitt_et_al_2010}. For instance, a study might examine the
question, ``To what extent do different teaching methods (\(T\))
influence students' conceptual understanding of a topic (\(Y\))?'' To
investigate this, the study could randomly assign students to two
groups, each using a different teaching method \((T_{i}=\{1,2\})\).
Students' conceptual understanding of the topic could be evaluated
through pairwise comparisons, resulting in a dichotomous outcome
\((Y_{i}=\{0,1\})\), indicating which student among those compared has a
higher level of understanding. The research question could be then
framed as the estimand, ``\emph{On average}, is there a difference in
conceptual understanding of the topic between the two groups of
students?'' This estimand could be mathematically expressed by the
associational quantity \(E[Y_{i}| T_{i}=1] - E[Y_{i}| T_{i}=2]\), where
\(E[\cdot]\) denotes the expected value. An example of this approach is
seen in \citet{Jones_et_al_2019}.

Researchers then proceed to identify the estimands.
\emph{Identification} refers to the process of accurately computing an
estimand using an estimator. An \emph{estimator} is a method or function
that maps data into an estimate \citep{Neal_2020}. \emph{Estimates} are
numerical values that approximate the estimand and are derived through
\emph{estimation}, which refer to the process of integrating data with
an estimator \citep{Everitt_et_al_2010}. Although various methods can
approximate an estimand, researchers prioritize estimators with
desirable properties that ensure the accuracy of estimates. For
instance, the BTL model \citep{Bradley_et_al_1952, Luce_1959} is an
estimator known for its effectiveness in modeling pairwise comparisons,
yielding accurate estimates for the target of inference discussed
earlier, as long as the assumptions of the model are met.

However, many studies aim to understand the mechanisms underlying
specific data types and also seek to establish causal relationships
rather than merely infer associations. In the earlier example, the
differences between groups obtained using the BTL model, referred to as
the associational estimates, can be interpreted as causal because the
data were collected through a randomized experiment. Randomized
experiments enable the causal interpretation of associational estimates
by ensuring several key properties, including common support, no
interference, and consistency \citep{Morgan_et_al_2014, Neal_2020}. The
most crucial property, however, is that randomization effectively
eliminates confounding. \emph{Confounding} occurs when an external
variable influences both the outcome and the variable of interest,
leading to spurious associations \citep{Everitt_et_al_2010}.
Randomization mitigates this issue by decoupling the intervention
assignment mechanism, such as assigning students to different groups,
from other variables and outcomes \citep{Morgan_et_al_2014, Neal_2020}.

@@

While a detailed explanation of causes and counterfactuals is beyond the
scope of this document, it is important to note that causal inference
addresses the \emph{fundamental problem of causality} through the use of
counterfactuals \citep{Neal_2020}. The framework defines
\emph{individual causal effects} (ICE) as the difference between an
individual's observed and unobserved potential outcomes. These
unobserved potential outcomes, known as \emph{counterfactuals},
represent hypothetical scenarios that are \emph{contrary to fact}, where
alternative outcomes for an individual, resulting from a specific cause,
are neither observed nor observable
\citep{Neal_2020, Counterfactual_2024}. The framework then extends this
estimand to \emph{average causal effects} (ACE), which captures the
difference between sample averages of observed potential outcomes and
counterfactuals. Finally, much like in an experiment, the framework
identifies the ACE from associational estimates by conditioning for a
\emph{sufficient adjustment set} of variables, ensuring the absence of
confounding \citep{Morgan_et_al_2014}.

@@

The motivating example in Section~\ref{sec-framework-structural} can
further illustrate how to use the five fundamental building blocks to
construct a system's causal structure. In this scenario, the
investigator aims to determine whether, \emph{on average}, there is a
difference in conceptual understanding of a topic \((Y_{i}=\{0,1\})\)
between two groups of students \((T_{i}=\{1,2\})\), described by the
estimand \(E[Y_{i}| T_{i}=1] - E[Y_{i}| T_{i}=2]\). However, unlike the
previous case, an experiment cannot be conducted, and the problem
suggests that the country to which a student belongs (\(X\)) may
influence both \(T\) and \(Y\). Such scenarios are plausible, especially
when the teaching methods depend on software or access to technology,
which may be limited in certain countries {(maybe an example with more
impact?)}. Figure~\ref{fig-example} illustrates the plausible causal
structure of the motivating example. A detailed examination of Figures
\ref{fig-mdag_example1}, \ref{fig-sdag_example1}, and SCM
\ref{fig-scm_example1} reveals the presence of at least four of the five
fundamental building blocks. The figures display multiple descendants,
as indicated by pairwise relations such as \(X \rightarrow T\),
\(X \rightarrow Y\), and \(T \rightarrow Y\). Additionally, the figures
features multiple pairs of unconnected nodes, evident from the relations
\(e_{T} \perp\!\!\!\perp e_{X}\), \(e_{T} \perp\!\!\!\perp e_{Y}\), and
\(e_{X} \perp\!\!\!\perp e_{Y}\). Finally, the figures illustrate the
fork \(X \rightarrow \{T,Y\}\), and two colliders with
\(\{X,e_{T}\} \rightarrow T\) and \(\{X,T,e_{Y}\} \rightarrow Y\).

\includegraphics[width=1\textwidth,height=\textheight]{images/figures/mdag_example1.png}

\subcaption{\label{fig-mdag_example1}Magnified representation}

\includegraphics[width=0.55\textwidth,height=\textheight]{images/figures/sdag_example1.png}

\subcaption{\label{fig-sdag_example1}Standard representation}

\[
\begin{aligned}
  X & := f_{X}(e_{X}) \\
  T & := f_{Z}(X,e_{T}) \\
  Y & := f_{Y}(T,X,e_{Y}) \\
  e_{T} & \perp\!\!\!\perp e_{X} \\
  e_{T} & \perp\!\!\!\perp e_{Y} \\
  e_{X} & \perp\!\!\!\perp e_{Y}
\end{aligned}
\]

\subcaption{\label{fig-scm_example1}Structural causal model}

\caption{\label{fig-example}DAGs for a plausible causal structure in a
system.}

@@
